\documentclass[11pt]{article}

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[total={7in,9in}]{geometry}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{lmodern}
\usepackage{hyperref}

\title{Berkeley ZK MOOC homework assignment}
\author{ \\ Morgan Thomas \\ Casper Association \\ morgan@casper.network }


\begin{document}

\maketitle

\section{Question 1}

\subsection{Part (a)}

\textbf{Question.} Give an expression for the multilinear extension
$\widetilde{\mathsf{eq}}_\ell : \{0,1\}^\ell \times \{0,1\}^\ell \to \{0,1\}$, that is defined
as follows:

\begin{equation}
  \mathsf{eq}_\ell(a, b) = \begin{cases}
	  1 & \text{if}\ a = b \\
	  0 & \text{otherwise}
  \end{cases}
\end{equation}

Argue that this polynomial can be evaluated at any point in time $O(\ell)$.

\textbf{Answer.} Let:

\begin{equation}
	\widetilde{\mathsf{eq}}_\ell(a,b) = \prod_{i=1}^\ell 1 - a_i + b_i.
\end{equation}

This polynomial agrees with $\mathsf{eq}$ on all
$(a,b) \in \{0,1\}^\ell \times \{0,1\}^\ell$. It is also multilinear,
as I can see by induction on $\ell$. When $\ell = 1$, it is clearly
multilinear. When $\ell > 1$, I can observe that

\begin{equation}
	\widetilde{\mathsf{eq}}_\ell(a,b) = (1 - a_\ell + b_\ell) \cdot \widetilde{\mathsf{eq}}_{\ell-1} = \widetilde{\mathsf{eq}}_{\ell-1}(a,b) - a_\ell \cdot \widetilde{\mathsf{eq}}_{\ell-1}(a,b) + b_\ell \cdot \widetilde{\mathsf{eq}}_{\ell-1}(a,b),
\end{equation}

and $\widetilde{\mathsf{eq}}_{\ell-1}$ does not refer to $a_\ell$ or $b_\ell$, so
therefore $\widetilde{\mathsf{eq}}_\ell$ is multilinear if $\widetilde{\mathsf{eq}}_{\ell-1}$ is.

These polynomials can be evaluated using $\ell$ additive inverse operations (one on each $a_i$), $2\ell$ addition operations, and $\ell-1$ multiplication operations, so, they can be evaluated using $O(\ell)$ operations. Said operations are $O(1)$ time, so overall these polynomials can be evaluated in $O(\ell)$ time.

\subsection{Part (b)}

\textbf{Question.} Assume that $n$ is a power of two. Give an expression for $\widetilde{\mathsf{mult}}$ that can be evaluated at any point in time $O(\log n)$. The multiplication layer consists of $n = 2^{d-1}$ multiplication gates, where the $j$-th multiplication gate at layer $d - 1$ has both in-neighbors equal to the $j$-th input gate at layer $d$.

\textbf{Answer.} For all positive integers $\ell$ and bit vectors $a,b,c$ of length $\ell$, let:

\begin{equation}
	\mathsf{eq}^3_\ell(a, b, c) =
	\begin{cases}
		1 & \text{if}\ a = b = c, \\
		0 & \text{otherwise}.
	\end{cases}
\end{equation}

Define the multilinear extension $\widetilde{\mathsf{eq}}^3_\ell$ of $\mathsf{eq}^3_\ell$ as follows:

\begin{equation}
	\widetilde{\mathsf{eq}}^3_\ell(a, b, c) =
	\prod_{i=1}^\ell (a_i \cdot b_i \cdot c_i) + ((1 - a_i) \cdot (1 - b_i) \cdot (1 - c_i)).
\end{equation}

It should be easy to check that this is an extension of $\mathsf{eq}^3_\ell$.
It is also multilinear because each term of the big product is multilinear and
no two terms refer to the same variable (and this reasoning can be further done out
as a proof by induction as in the previous answer). Also observe that this expression
can be evaluated in $O(\ell)$ time.

Let $n$ be the number of multiplication gates. Let $d$ be the number of layers
of the circuit. Consider that the number of nodes in the circuit is $2^d-1+n$,
where $2^d-1$ is the number of nodes in a perfect binary tree of height $d-1$,
and $n$ is the number of input nodes. Consdering that $n = 2^{d-1}$, it requires
$d+1$ bits to represent an index into the list of nodes.

Let the nodes be indexed from left to right, starting from the bottom layer. 
Let the bits in the index bit strings be indexed in LSB first notation.
Let $k$ be the number of bits
in $n$. Note that $k = O(\log n)$ and $d+1-n = O(d) = O(\log n)$.

Let:

\begin{equation}
	\begin{array}{rcl}
		\widetilde{\mathsf{mult}}(a, b, c) &=& 
		\widetilde{\mathsf{eq}}^3_k([a_1, ..., a_n], [b_1, ..., b_n], [c_1, ..., c_n]) \\
		&\cdot& \widetilde{\mathsf{eq}}^3_{d+1-n}([b_{n+1}, ..., b_{d+1}], [c_{n+1}, ..., c_{d+1}], [0, ..., 0]) \\
		&\cdot& \widetilde{\mathsf{eq}}_{d+1-n}([a_{n+1}, ..., a_{d+1}], [1, 0, ..., 0]).
	\end{array}
\end{equation}

Observe that this defines the multilinear extension of $\mathsf{mult}(a, b, c)$,
and it can be evaluated in $O(\log n)$ since each of the subscripts on the
equality predicates is $O(\log n)$.

\subsection{Part (c)}

\paragraph{Question.} Assume that $n$ is a power of 2. Give an expression for $\widetilde{\mathsf{add}}$
that can be evaluated at any point in time $O(\log n)$. The addition layer $i$ consists of $2^i$
addition gates, where for $j \in \{0, 1, ..., 2^i - 1\}$, the $j$-th addition gate at layer $i$
has as its in-neighbors gates $2j$ and $2j+1$ at layer $i+1$.

\paragraph{Answer.} Let the nodes be indexed from left to right, starting from the bottom layer
and moving up. Let the indices be represented as bit strings in LSB first notation. Let $k$ be the
number of bits in $n$. Let $d$ be the number of layers in the circuit. As in part (b), the number of
nodes in the circuit is $2^d-1+n$, including $n$ input nodes, where $2^d-1$, the number of nodes in
a perfect binary tree of height $d-1$, is the number of non-input nodes. $k+2$ is the number of bits
in an index into the nodes.

Let $\mathsf{idx}(i, j)$ denote the index of the $j$-th (zero indexed from the left) node in the
$i$-th (zero indexed from the bottom) layer. Concretely,

\begin{equation}
	\begin{array}{rcl}
		\mathsf{idx}(i, j) &=& (2^d-1) - (2^{d-(i-1)}-1) + n + j \\
		&=& 2^d - 2^{d-i+1} + n + j.
	\end{array}
\end{equation}
This formula works because $2^d-1$ is the number of nodes in a perfect binary tree of height $d-1$,
and $2^{d-(i-1)}-1$ is the number of nodes in a perfect binary tree of height $d-1-(i-1)$, which is
in other words the number of nodes above layer $i$ in the circuit. $n$ is the number of nodes in
the bottom layer, and $j$ is the number of nodes in layer $i$ before the node with index
$\mathsf{idx}(i, j)$.

A sum of multilinear polynomials is a multilinear polynomial. Therefore, to compute the multilinear
extension of the wiring predicate $\mathsf{add}$, it suffices to compute the multilinear extension
of the addition wiring predicate for each layer, that is, for all $2 \leq i < d$, the multilinear
extension of

\begin{equation}
	\mathsf{add}_i(a, b, c) =
	\begin{cases}
		1 & \exists j \in [0,2^{d-i}), a = \mathsf{idx}(i, j) \wedge b = \mathsf{idx}(i-1, 2j) \wedge c = \mathsf{idx}(i-1, 2j+1), \\
		0 & \text{otherwise}.
	\end{cases}
\end{equation}

Let these layer-specific wiring predicates be broken down as a product of two predicates, one
stating that an index belongs to a particular layer, and one stating that $b_j = 2a_j$ and $c_j = 2a_j + 1$,
where $a_j$ is the $j$ of $a$, and so forth:

\begin{equation}
	\mathsf{layer}_i(a) =
	\begin{cases}
		1 & \exists j \in [0,2^{d-i}), a = \mathsf{idx}(i, j), \\
		0 & \text{otherwise}.
	\end{cases}
\end{equation}

\begin{equation}
	\mathsf{mask}_i(a) = a \text{\&} (2^{d-i-1}-1).
\end{equation}

Here \& denotes the bitwise AND operation. This bit mask is supposed to extract $j$;
that is, $\mathsf{mask}_i(a) = a_j$ if $a$ is in layer $i$.

\begin{equation}
	\mathsf{lidx}_i(a, b, c) =
	\begin{cases}
		1 & \mathsf{mask}_{i-1}(b) = 2 \cdot \mathsf{mask}_i(a) \wedge \mathsf{mask}_{i-1}(c) = 2 \cdot \mathsf{mask}_i(a) + 1, \\
		0 & \text{otherwise}.
	\end{cases}
\end{equation}

\begin{equation}
	\mathsf{add}_i(a, b, c) =
	\mathsf{layer}_i(a) \cdot \mathsf{layer}_{i-1}(b) \cdot \mathsf{layer}_{i-1}(c) \cdot \mathsf{lidx}_i(a, b, c).
\end{equation}

The purpose of decomposing $\mathsf{add}_i$ in this way is that the four terms of the product are each
dependent on disjoint subsets of the bits of $a, b,$ and $c$. So, to find the multilinear extension of
$\mathsf{add}_i$, it suffices to find the multilinear extensions of each of the four terms of the product.
Let's write the multilinear extensions explicitly, which will show the claim that the terms are dependent
on disjoint sets of bits and therefore that their product is multilinear:

\begin{equation}
	\widetilde{\mathsf{layer}}_i(a) = \widetilde{\mathsf{eq}}_{i+1}([a_{k+1}, a_k, a_{k-1}, ..., a_{k+1-i}], [1, ..., 1, 0]).
\end{equation}

\begin{equation}
	\begin{array}{rcl}
		\widetilde{\mathsf{lidx}_i}(a) &=& \widetilde{\mathsf{eq}}_1([b_0], [0]) \cdot \widetilde{\mathsf{eq}}_1([c_0, [1]) \\
		&\cdot& \widetilde{\mathsf{eq}}^3_{k+1-i}([a_0, ..., a_{k-i}], [b_1, ..., b_{k+1-i}], [c_1, ..., c_{k+1-i}]).
	\end{array}
\end{equation}

The remaining definitions fall out of what has been said so far:

\begin{equation}
	\widetilde{\mathsf{add}}_i(a, b, c) = \widetilde{\mathsf{layer}}_i(a) \cdot \widetilde{\mathsf{layer}}_{i-1}(b) \cdot \widetilde{\mathsf{layer}}_{i-1}(c) \cdot \widetilde{\mathsf{lidx}}_i(a, b, c).
\end{equation}

\begin{equation}
	\widetilde{\mathsf{add}}(a, b, c) = \sum_{i=1}^{d-1} \widetilde{\mathsf{add}}_i(a, b, c).
\end{equation}

TODO: argue evaluation time

\section{Question 2}

This question asks how to implement PLONKish arithmetization, with custom gates. The assumption of this
answer is that this is equivalent to Halo 2's PLONKish arithmetization without fixed columns or
lookup arguments. It features custom gates which can use variables with relative row offsets, with
no limit on the magnitude of the offsets. These custom gates are defined as polynomials constrained to equal
zero at all rows. It also features equality constraints, defined as a permutation $W$ of the cells (i.e.,
a rotation of $\Omega$), with the equality constraints stating that $T(W(i)) = T(i)$, for all $i \in \Omega$,
where $T$ is the trace polynomial.

\subsection{Part (a)}

$\Omega$ is the multiplicative subgroup of $\mathbb{F}_p$ generated by $\omega$, which is a $(d-1)$th root of unity.
$d$ is $(n+1)|C| + |I|$, where $n$ is the number of inputs per gate, $C$ is the set of gates in
the arithmetic circuit, and $I$ is the set of inputs
to the circuit (both instance and witness inputs). The significance of $\Omega$ is that the trace polynomial
$T$ encodes the values in the trace (the inputs, intermediate values, and outputs) along the points in $\Omega$.

$\Omega_{\text{inp}}$ is the set of values in $\Omega$ which are the coordinates of the instance inputs: specifically,
\begin{equation}
	\Omega_{\text{inp}} = \omega^{-1}, ..., \omega^{-|I_x|},
\end{equation}
where $I_x$ is the set of instance inputs.

$\Omega_{\text{gates}}$ is the set of values in $\Omega$ which are the coordinates of the outputs of the gates:
specifically,
\begin{equation}
	\Omega_{\text{gates}} = \{\omega^{(n+1)i+n}\}_{i=0}^{|C|-i}.
\end{equation}

The trace polynomial $T$ is a polynomial function which contains, along the points in $\Omega$, the input and
output values of each gate in the circuit, as well as the inputs of the circuit. $T$ is obtained by
interpolating over the following set of input-output pairs:
\begin{equation}
	\{(\omega^{-i}, v(i))\}_{i=0}^{|I|-1} \cup \left( \bigcup_{j=0}^{n-1} \{(\omega^{(n+1)i+j}, g_j(i)\}_{i=0}^{|C|-1} \right )
	\cup \{\omega^{(n+1)i+n}, h(i)\}_{i=0}^{|C|-1}.
\end{equation}
Here, $v$ denotes the function mapping the index of an input to its value. $g_j$ denotes the function mapping the
index of a gate to its $j$th input value. $h$ denotes the function mapping the index of a gate to its output value.

Let's use, instead of one selector polynomial, a collection of selector polynomials, one for each \emph{type}\/
of gate. Let $m$ be the number of gate types. For each $i \in [0,m)$, the selector polynomial $S_i$ is a
polynomial function which contains, along the points in $\Omega_{\text{gates}}$, the value 1 for each gate
of type $i$, and the value 0 for each gate not of type $i$. $S_i$ is obtained by interpolation over the
set of input-output pairs specified in the last sentence.

\subsection{Part (b)}

The IOP needs to prove, in summary, the following claims:

\begin{enumerate}
	\item The trace polynomial encodes the correct statement (i.e., instance inputs);
	\item Every gate is evaluated correctly;
	\item The wiring / equality constraint is satisfied;
	\item The output of each gate is zero.
\end{enumerate}

To prove (1), use a zero test to check that $T(\omega^{-i}) = v(i)$ for all $i \in [0,|I_x|)$.
The zero test needs to check that a certain polynomial is zero over a point set, so,
let $\widetilde{v}$ be the polynomial which can be computed publicly by interpolating the
following point set:
\begin{equation}
	\{(\omega^{-i}, v(i)\}_{i=0}^{|I_x|-1}.
\end{equation}
The zero test checks that the polynomial $T(y) - \widetilde{v}(y) = 0$ for all $i \in \Omega_{\text{inp}}$.
The size of the set the test is used in on is $|I_x|$, the number of instance inputs.
The degree of the polynomial is $|\Omega|$.

To prove (2), use a zero test to check that the following polynomial is zero for all $y \in \Omega_{\text{gates}}$:
\begin{equation}
	T(y) - \sum_{j=0}^{m-1} S_j(y) \cdot G_i(T, y).
\end{equation}
Here $G_i$ denotes the definition of the $i$th custom gate, which is parameterized by the trace polynomial and
the output coordinate. The degree of this polynomial is no less than $|\Omega|$, since $|\Omega|$ is the degree
of $T$, but there is no upper bound on the degree of the polynomial as it depends on the degrees of the $G_i$.
This test is applied on a set of size $|C|$, the number of gates.

To prove (3), use a prescribed permutation check to show that for all $y \in \Omega$, $T(y) = T(W(y))$.
The prescribed permutation check reduces to a product check on a polynomial of degree $2|\Omega|$
on a set of size $|\Omega|$.

To prove (4), use a zero test to show that $T(y) = 0$ for all $y \in \Omega_{\text{gates}}$. 
The polynomial $T$ is of degree $|\Omega|$ and it is used on a set of size $|C|$, the number of gates.

\end{document}
